---
title: "Mouse vendors data analysis"
output: html_document
---

```{r}
library(ggtree)
library(lmerTest)
library(tidyverse)

# Other packages used but not loaded into global namespace:
# For loading data
# - taxafmt (GitHub)
# For taxonomic overview
# - pheatbuilder (GitHub)
# - ggsci
# For alpha/beta diversity measures
# - vegan
# - abdiv (GitHub)
# For distance matrix work
# - usedist (GitHub)
# - adonisplus (GitHub)
# For plotting
# - ggbeeswarm
# For statistical tests
# - broom
# - broom.mixed
# For supervised learning
# - pROC
# - MASS
# - randomForest
# - randomForestExplainer
```

## Colors

```{r}
vendor_colors <- c(
  "Charles River" = "#484349",
  "Envigo" = "#55778B",
  "Jackson" = "#109648",
  "Taconic" = "#F2A65A")
sample_type_colors <- c(
  "Feces" = "#484349",
  "Cecal contents" = "#c6c2c7"
)
```

## Data import

### Metadata

```{r}
s <- read_tsv("vendors-data/vendor-samples.tsv") |>
  select(
    sample_id = `#SampleID`, sample_type = SampleType,
    subject_id = Mouse_Number, vendor = Mouse_Source_Vendor) |>
  mutate(sample_type = fct_recode(
    sample_type,
    "Feces" = "Stool",
    "Cecal contents" = "Cecal Content")) |>
  mutate(sample_type = fct_relevel(sample_type, "Feces")) |>
  mutate(subject_id = paste("Mouse", subject_id)) |>
  mutate(vendor = fct_recode(vendor, "Jackson" = "JAX")) |>
  mutate(vendor_label = str_wrap(vendor, width = 7))
```

### Taxa counts

```{r}
cts <- read_tsv("vendors-data/feature-table.tsv", skip = 1) |>
  rename(otu_id = `#OTU ID`) |>
  pivot_longer(
    -otu_id,
    names_to = "sample_id",
    values_to = "reads")

taxa <- read_tsv("vendors-data/taxonomy.tsv") |>
  select(otu_id = `Feature ID`, lineage = Taxon) |>
  mutate(taxafmt::split_lineage(lineage)) |>
  mutate(across(Kingdom:Species, taxafmt::remove_rank_prefix)) |>
  mutate(Species = taxafmt::make_binomial_name(Genus, Species)) |>
  mutate(assignment = taxafmt::format_taxa(pick(Kingdom:Genus))) |>
  filter(!str_detect(Class, "Chloroplast")) |>
  filter(!str_detect(Family, "mitochondria")) |>
  filter(!is.na(Phylum))

cts <- cts |>
  filter(otu_id %in% taxa$otu_id)
```

```{r}
props <- cts |>
  group_by(sample_id) |>
  mutate(prop = reads / sum(reads)) |>
  ungroup() |>
  left_join(taxa, join_by(otu_id)) |>
  group_by(sample_id, assignment) |>
  summarise(prop = sum(prop), .groups = "drop")
```

### Alpha diversity

```{r}
faith_pd <- read_tsv("vendors-data/alpha-diversity-faithpd.tsv") |>
  rename(sample_id = 1)
```

```{r warning=FALSE}
adiv <- cts |>
  group_by(sample_id) |>
  summarise(
    shannon = abdiv::shannon(reads),
    simpson = abdiv::simpson(reads),
    richness = vegan::rarefy(reads, 1000)) |>
  left_join(faith_pd, join_by(sample_id))
```

### Beta diversity

```{r}
wu <- read.table("vendors-data/distance-matrix-wu.tsv") |>
  as.dist()
```

```{r}
uu <- read.table("vendors-data/distance-matrix-uu.tsv") |>
  as.dist()
```

```{r}
bc <- cts |>
  usedist::pivot_to_matrix(sample_id, otu_id, reads) |>
  vegan::vegdist(method = "bray")
```

```{r}
jacc <- cts |>
  usedist::pivot_to_matrix(sample_id, otu_id, reads) |>
  vegan::vegdist(method = "jaccard", binary = TRUE)
```

## Taxonomic overview

### Heatmap

```{r fig.width=10, fig.height=5}
s_annotation <- s |>
  select(sample_id, vendor, sample_type) |>
  arrange(vendor, sample_type)
props |>
  group_by(assignment) |>
  filter(mean(prop) > 0.001) |>
  pivot_wider(names_from = sample_id, values_from = prop, values_fill = 0) |>
  select(assignment, all_of(s_annotation$sample_id)) |>
  pheatbuilder::pheat() |>
  pheatbuilder::pheat_annotate_cols(s_annotation) |>
  pheatbuilder::pheat_display_cols(gaps = c(8, 16, 24, 32)) |>
  pheatbuilder::pheat_color_saturated(saturation_limit = 0.6) |>
  pheatbuilder::pheat_annotation_color(
    vendor = vendor_colors, sample_type = sample_type_colors)
```

### Stacked bars

```{r fig.width=8, fig.height=5}
top_assignments <- props |>
  group_by(assignment) |>
  summarise(mean_prop = mean(prop), .groups = "drop") |>
  filter(mean_prop > 0.02)

props |>
  mutate(assignment = if_else(
    assignment %in% top_assignments$assignment, assignment, NA_character_)) |>
  mutate(assignment = fct_explicit_na(assignment, "Other")) |>
  group_by(sample_id, assignment) |>
  summarise(prop = sum(prop), .groups = "drop") |>
  left_join(s, join_by(sample_id)) |>
  ggplot(aes(x = subject_id, y = prop, fill = assignment)) +
  geom_col() +
  facet_grid(sample_type ~ vendor, scales = "free_x") +
  ggsci::scale_fill_frontiers() +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "", y = "Relative abundance", fill = "") +
  theme_bw() +
  theme(axis.text.x = element_blank())
```

## Alpha diversity

### Plot

```{r}
s |>
  left_join(adiv, join_by(sample_id)) |>
  ggplot(aes(x = vendor_label, y = shannon, color = vendor)) +
  geom_boxplot(outlier.shape = NA) +
  ggbeeswarm::geom_quasirandom(aes(shape = sample_type)) +
  scale_color_manual(values = vendor_colors, guide = "none") +
  scale_shape_manual(values = c(19, 1)) +
  scale_y_continuous(limits = c(0, 4.5)) +
  labs(x = "Vendor", y = "Shannon diversity", shape = "") +
  theme_bw()
ggsave("figures/shannon_diversity.pdf", width = 4, height = 2.6, useDingbats = F)
```

### Test

Simple linear model, two-group comparison

```{r}
s |>
  filter(vendor %in% c("Envigo", "Jackson"), sample_type %in% "Feces") |>
  left_join(adiv, join_by(sample_id)) |>
  lm(shannon ~ vendor, data = _) |>
  summary()
```

Equivalent to a t-test

```{r}
s |>
  filter(vendor %in% c("Envigo", "Jackson"), sample_type %in% "Feces") |>
  left_join(adiv, join_by(sample_id)) |>
  t.test(shannon ~ vendor, data = _, var.equal = TRUE)
```

Linear model to compare multiple groups against a reference group

```{r}
s |>
  filter(sample_type %in% "Feces") |>
  left_join(adiv, join_by(sample_id)) |>
  lm(shannon ~ vendor, data = _) |>
  summary()
```

Mixed-effects model to compare repeated measurements from the same mouse

```{r}
s |>
  filter(vendor %in% c("Envigo", "Jackson")) |>
  left_join(adiv, join_by(sample_id)) |>
  lmer(shannon ~ vendor * sample_type + (1|subject_id), data = _) |>
  summary()
```

## Beta diversity

### Plot

```{r}
s |>
  adonisplus::pcoaplus(uu, sample_id_var = sample_id) |>
  plot(color = vendor, shape = sample_type) +
  facet_grid(~ sample_type) +
  scale_shape_manual(values = c(19, 1)) +
  scale_color_manual(values = vendor_colors) +
  theme_bw()
```

### Test

```{r}
s |>
  adonisplus::adonisplus(
    jacc, distmat ~ vendor * sample_type, 
    rep_meas_var = subject_id, sample_id_var = sample_id,
    shuffle = c(sample_type = "within", vendor = "between"))
```

```{r}
s |>
  filter(sample_type %in% "Feces") |>
  adonisplus::adonispost(
    jacc, distmat ~ vendor, which = vendor,
    sample_id_var = sample_id, permutations = 999)
```

```{r}
2 / choose(8, 4)
```

## Taxonomic abundance

### Plot

```{r}
props_toTest <- props |>
  group_by(assignment) |>
  filter(mean(prop) > 0.01) |>
  ungroup()
```

```{r}
props_toTest |>
  left_join(s, join_by(sample_id)) |>
  mutate(prop = prop + 1e-6) |>
  mutate(assignment = str_replace(assignment, " - ", "\n")) |>
  mutate(assignment = str_replace(assignment, "unclassified", "uncl.")) |>
  ggplot(aes(x = vendor, y = prop, shape = sample_type)) +
  ggbeeswarm::geom_quasirandom() +
  facet_wrap(~ assignment, scales = "free_y") +
  scale_y_log10() +
  scale_shape_manual(values = c(19, 1)) +
  labs(x = "", y = "Relative abundance") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

### Test

```{r}
props_toTest |>
  left_join(s, join_by(sample_id)) |>
  filter(sample_type %in% "Feces") |>
  mutate(logprop = log10(prop + 1e-6)) |>
  nest_by(assignment) |>
  reframe(broom::tidy(lm(logprop ~ vendor, data = data))) |>
  filter(!(term %in% "(Intercept)")) |>
  group_by(term) |>
  mutate(fdr = p.adjust(p.value, method = "fdr")) |>
  mutate(sig = cut(fdr, c(-Inf, 0.001, 0.01, 0.05, Inf), c("***", "**", "*", "")))
```

```{r}
props_toTest |>
  left_join(s, join_by(sample_id)) |>
  filter(sample_type %in% "Feces") |>
  mutate(vendor = fct_relevel(vendor, "Jackson")) |> ### Compare to Jackson
  mutate(logprop = log10(prop + 1e-6)) |>
  nest_by(assignment) |>
  reframe(broom::tidy(lm(logprop ~ vendor, data = data))) |>
  filter(!(term %in% "(Intercept)")) |>
  group_by(term) |>
  mutate(fdr = p.adjust(p.value, method = "fdr")) |>
  mutate(sig = cut(fdr, c(-Inf, 0.001, 0.01, 0.05, Inf), c("***", "**", "*", "")))
```

```{r}
props_toTest |>
  left_join(s, join_by(sample_id)) |>
  mutate(logprop = log10(prop + 1e-6)) |>
  nest_by(assignment) |>
  reframe(broom.mixed::tidy(lmerTest::lmer(
    logprop ~ vendor * sample_type + (1|subject_id), data = data))) |>
  filter(!(term %in% "(Intercept)")) |>
  filter(effect %in% "fixed") |>
  select(-effect, -group, -statistic, -df) |>
  group_by(term) |>
  mutate(fdr = p.adjust(p.value, method = "fdr")) |>
  mutate(sig = cut(fdr, c(-Inf, 0.001, 0.01, 0.05, Inf), c("***", "**", "*", "")))
```

## Supervised learning

```{r}
props_wide <- props |>
  left_join(s, join_by(sample_id)) |>
  filter(vendor %in% c("Charles River", "Envigo")) |>
  mutate(vendor = fct_drop(vendor)) |>
  group_by(assignment) |>
  filter(mean(prop > 0) > 0.5) |>
  ungroup() |>
  mutate(assignment = fct_reorder(assignment, desc(prop))) |>
  pivot_wider(
    id_cols = c(sample_id, vendor), 
    names_from = assignment, values_from = prop)
```

### Logistic recgression

```{r}
logistic_taxa <- c(
  "Bacteroidetes - unclassified Rikenellaceae",
  "Firmicutes - Oscillospira")
```


```{r}
logistic_model <- props_wide |>
  select(vendor, all_of(logistic_taxa)) |>
  glm(vendor ~ ., family = "binomial", data = _)
logistic_model |>
  summary()
```

```{r}
predict(logistic_model)
logistic_vendor <- factor(
  predict(logistic_model) > 0,
  labels = c("Charles River", "Envigo"))
logistic_vendor
```

```{r}
table(Actual = props_wide$vendor, Predicted = logistic_vendor)
7 / 8 # True positive rate
1 / 8 # False positive rate
```

```{r}
logistic_confusion <- function (threshold) {
  actual <- props_wide$vendor %in% "Envigo"
  predicted <- predict(logistic_model) > threshold
  tibble(
    tpr = sum(predicted & actual) / sum(actual),
    fpr = sum(predicted & (!actual)) / sum(!actual))
}
logistic_roc_data <- tibble(threshold = seq(-20, 5, 0.5)) |>
  group_by(threshold) |>
  reframe(logistic_confusion(threshold))
```

```{r}
logistic_roc_data |>
  pivot_longer(-threshold) |>
  ggplot(aes(x = threshold, y = value, color = name)) +
  geom_step()
```

```{r}
logistic_roc_data |>
  arrange(tpr, fpr) |>
  ggplot(aes(x = fpr, y = tpr)) +
  geom_step() +
  coord_equal() +
  scale_x_continuous(limits = c(0, 1)) +
  theme_bw()
```

```{r}
logistic_roc <- pROC::roc(
  props_wide$vendor %in% "Envigo",
  predict(logistic_model))
pROC::auc(logistic_roc)
62 / 64
```

```{r}
pROC::ggroc(logistic_roc) +
  coord_equal() +
  theme_bw()
```

### Linear discriminant analysis (LDA)

```{r}
lda_taxa <- c(
  "Bacteroidetes - unclassified Rikenellaceae",
  "Firmicutes - Oscillospira",
  "Firmicutes - unclassified Ruminococcaceae",
  "Verrucomicrobia - Akkermansia")
```


```{r}
lda_model <- props_wide |>
  select(-sample_id) |>
  select(vendor, all_of(lda_taxa)) |>
  MASS::lda(vendor ~ ., data = _)
```

```{r}
predict(lda_model)
lda_values <- predict(lda_model)$x |>
  as.data.frame() |>
  mutate(sample_id = props_wide$sample_id)
```

```{r}
table(Actual = props_wide$vendor, Predicted = predict(lda_model)$class)
```

```{r}
props_wide |>
  bind_cols(predict(lda_model)$x) |>
  ggplot(aes(x = LD1, y = sample_id, color = vendor)) +
  geom_point() +
  geom_vline(xintercept = 0, linetype = "dashed") +
  theme_bw()
```

```{r}
lda_roc <- pROC::roc(
  props_wide$vendor %in% "Envigo",
  predict(lda_model)$x[,1])
pROC::auc(lda_roc)
plot(lda_roc)
```

```{r}
lda_model$scaling |>
  as.data.frame() |>
  rownames_to_column("assignment") |>
  mutate(assignment = fct_reorder(assignment, LD1)) |>
  ggplot(aes(x = LD1, y = assignment)) +
  geom_point() +
  geom_vline(xintercept = 0, linetype = "dashed") +
  theme_bw()
```

### Random forests

```{r}
rforest_predictors <- props_wide |>
  select(-sample_id, -vendor)
rforest_response <- props_wide$vendor
rforest_model <- randomForest::randomForest(
  x = rforest_predictors, y = rforest_response)
rforest_model
```

```{r}
predict(rforest_model)
```


```{r}
randomForest::varImpPlot(rforest_model)
```

```{r}
randomForestExplainer::plot_min_depth_distribution(rforest_model)
```

